Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-09-01 ms.tex
finishing edits to discussion and summary

The solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere originally derived by \cite{2006ApJ...649...14D} (called $H_0$ here) has been generalized to allow a monochromatic source of photons with frequencies away from line center, and a correction to the incorrect $J=0$ surface boundary condition is made. This correction allows the proper condition $J=\sqrt{3}H$ to be satisfied, and comes from the introduction of the terms $J_{\rm bc}$ and $H_{\rm bc}$ which are solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. In practice, the continuous expansions are discretized and the Fourier coefficients solved for via a matrix equation. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$ where $a$ is the damping parameter and $\tau_0$ is the line-center optical depth. Thus, for very large optical depths, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower optical depths. Since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, these solutions are invalid in areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. The peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$; as such, calculations performed at low optical depths are likely to be inaccurate due to the close proximity of the spectral peak and the Doppler core of the line profile. By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions %and correctly tracks changes in the spectrum of escaping photons as the monochromatic source frequency $x_s$ is altered, 
provided a sufficiently large optical depth $\tau_0$. Specifically, this solution shows improvement over previous solutions that utilized a $J=0$ surface boundary condition, such as \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others. The surface boundary condition correction $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which suggests the previously cited solutions are too large at the spectral peaks and too small further out in the wing. This appears to be the consequence of not enforcing the correct frequency-dependent boundary conditions at the surface of the medium.

The primary application of this solution for resonance-line radiation transfer at large optical depths is to develop a computationally efficient random walk model in spherical geometry. There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations. The simplest class of methods that is specific to resonance line transfer are core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A}. These methods focus on Monte Carlo transfer in the wings of the line and avoid scatterings in the core, but

Previous work by \citet{2002ApJ...567..922A,2015MNRAS.449.4336S} discussed Monte Carlo code acceleration for optical depths as large as these in a uniform plane-parallel medium with no absorption, but these calculations would need to be generalized to spherical symmetry in order to construct a model of resonant scattering in exoplanet atmospheres as discussed in Section \ref{sec:intro}.

   Line core photons are pushed to the wings by biasing the atomic velocity distribution to high velocities so the opacity is lower and escape is more rapid.

A second class of methods are hybrid diffusion methods \citep{2001JCoPh.172..543G}, such as discrete diffusion Monte Carlo \citep{2007JCoPh.222..485D}. This involves designating sufficiently optically thick simulations cells for computation via a finite difference representation of the radiation transfer equation. This method is particularly attractive for gray opacity applications where a single diffusion equation can be computed.  For the resonant scattering problem, where the opacity is a strong function of frequency, implementation of this method is more complex and computationally demanding as it requires the solution of many frequency groups, but can lead to a considerable speed up \citep{2018MNRAS.479.2065S}.

A third class of methods are modified random walk methods \citep{1984JCoPh..54..508F,2009A&A...497..155M, 2010A&A...520A..70R}. In this case a spherical volume (usually the largest within a simulation cell) is chosen when the scattering mean-free-path is small compared with the zone size.  An outgoing photon is then randomly sampled on the surface of this outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions and escape times, based on solutions to the diffusion equation within a uniform sphere. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \cite{1990ApJ...350..216N} solution.

%The need for accurate distributions from which to draw outgoing photons' properties motivates this work; namely, an analytic solution to the steady-state problem of \lya resonant scattering in spherical geometry. In addition, a semi-analytic solution to time-dependent \lya diffusion through the same medium is explored to quantify how long the photons are expected to spend in the domain before escaping. These solutions will be shown to accurately represent the photon distributions in frequency and time-to-escape at the surface of a spherical volume of gas, highlighting their potential for use as a Monte Carlo accelerant in magnetohydrodynamic calculations of phenomena involving \lya resonance-line radiation.

Radiation forces due to \lya transfer have been calculated in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. In the grid cells of a radiation hydrodynamic simulation, it would be necessary to calculate the radiation force on each cell face due to \lya at line center optical depths of up to 1 million or more. A discussion of Monte Carlo code acceleration for optical depths as large as these in a uniform, plane-parallel medium with no absorption is presented in  \citet{2002ApJ...567..922A,2015MNRAS.449.4336S}. In terms of the time dependence, \citet{1994ApJ...427..603R} introduce an ansatz for the Fokker-Planck diffusion equation in order to quantify the relaxation timescale necessary to reach a quasi-static solution. Notably, the full Voigt function is used for the line profile rather than restricting scattering to the damping wing only.

\section{Summary and Conclusions}

% Move summary to discussion - currently written more like a discussion. 
% Write summary like it's a complement to the abstract. Cliffnotes version for people who don't want to read it.
% We re-examined previous solutions, noting that they use this incorrect boundary conditoin. We re-examined the problem with an appropriate bc, examine spectral and time dpeendence, key results are these, plan to use in a MC acceleration scheme in the future. Comment on comparison with MC. One, maybe two paragraphs.

The time-dependence of the problem is solved under a $J=0$ boundary condition to characterize the distribution of photon escape times. The differential equation cannot be solved except with a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency $x_s$, the flux at the surface of the sphere is found as a function of $t$ and $\nu$, expressed as a sum over many eigenmodes $n$ and $m$---spatial and frequency modes, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at high optical depths until they diffuse outward in frequency, biasing the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. When calculated as a cumulative distribution function using a rejection method, the fit can be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method. As will be shown in future work, these calculations can be implemented as an acceleration scheme for Monte Carlo resonance-line radiation transfer at optical depths that would nominally be computationally prohibitive.

>>>>

Models of heating and escape of hydrogen gas in exoplanet atmospheres can be constructed with an accurate treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but requires a solution of radiative transfer in a sphere and is limited by its high computational demand for large optical depths where there are many photon scatterings before escape. The primary goal of this work is to present a solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere. We have generalized a solution derived by \cite{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center, and we have made a correction to the incorrect $J=0$ surface boundary condition. This correction allows the proper condition $J=\sqrt{3}H$ to be satisfied, and comes from the introduction of the terms $J_{\rm bc}$ and $H_{\rm bc}$ which are solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. In practice, the continuous expansions are discretized and the Fourier coefficients solved for via a matrix equation. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$ where $a$ is the damping parameter and $\tau_0$ is the line-center optical depth. Thus, for very large optical depths, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower optical depths. Since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, these solutions are invalid in areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. The peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$; as such, calculations performed at low optical depths are likely to be inaccurate due to the close proximity of the spectral peak and the Doppler core of the line profile. By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions %and correctly tracks changes in the spectrum of escaping photons as the monochromatic source frequency $x_s$ is altered, 
provided a sufficiently large optical depth $\tau_0$. Specifically, this solution shows improvement over previous solutions that utilized a $J=0$ surface boundary condition, such as \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others. The surface boundary condition correction $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which suggests the previously cited solutions are too large at the spectral peaks and too small further out in the wing, where the error scales approximately as $(a\tau_0)^{-1/3}$. This appears to be the consequence of not enforcing the correct frequency-dependent boundary conditions at the surface of the medium.

Additionally, the time-dependence of the problem is solved under a $J=0$ boundary condition to characterize the distribution of photon escape times. The differential equation cannot be solved except with a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency $x_s$, the flux at the surface of the sphere is found as a function of $t$ and $\nu$, expressed as a sum over many eigenmodes $n$ and $m$---spatial and frequency modes, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to agree well with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the number of spatial terms included in the sum over eigenmodes, as the Doppler core can trap photons at high optical depths until they diffuse outward in frequency, biasing the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, a fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints. The cumulative distribution function could be calculated using a rejection method, which would then be used to sample photon escape times directly rather than computing their travel times by the Monte Carlo method.

This leads in to the primary application of this work: to develop a computationally efficient random walk model in spherical geometry. There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S}. Another approach with wide application is modified random walk methods, such as those discussed in \citet{2002ApJ...567..922A,2015MNRAS.449.4336S} in which Monte Carlo code acceleration is performed for large optical depths in a uniform plane-parallel medium with no absorption. These calculations would need to be generalized to spherical symmetry in order to construct a models of resonant scattering in exoplanet atmospheres as discussed in Section \ref{sec:intro}. The underlying principle is the same, however: an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \cite{1990ApJ...350..216N} solution, but this solution of course does not use the correct frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer on the faces of each grid cell in the simulation. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more. Such considerations will be taken in future work while developing this solution into a 1D hydrodynamic simulation using radiation forces from accelerated Monte Carlo \lya transfer.

\section{Summary}

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition as a frequency-dependent quantity, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a corrective boundary term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ diminishes at larger optical depths, following a predictable $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence back into the transfer equation leads to another solution which is obtained with a discretized eigenfunction expansion and solved numerically. We demonstrate that it agrees with the steady-state result when integrated over time, though its rate of numerical convergence is quite slow and requires a sum over many modes to become accurate. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of direct Monte Carlo simulations that would ordinarily be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of line-center optical depths, noting general agreement in the resulting escape time distributions. In future work, these distributions will be fitted and used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.
